MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-02 main/MSstatsPTM_Main.tex
Comments on the evaluations

\medskip \noindent {\bf Experimental Design:} Figure \ref{fig:benchmark-design}(a) overviews the experimental design. Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E. coli} lysate, and human spike-in ub-peptide mixture. Unmodified peptides from human lysate were viewed as the global proteome. Background {\it E. coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling between Mix 1 and 3, and between Mix 2 and 4.  50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides in a mixed background of lysates from {\it E. coli} and a human cell line. Quantitative changes in protein and site abundance of these 20 human proteins were the target of the benchmark. In particular, we distinguished the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed). The true log-fold changes between the relevant components of the relevant mixtures are summarized in Figure \ref{fig:benchmark-design}(b).  Two replicate mixtures were created per condition.

@@

The performance of the proposed method was evaluated on simulated and spike-in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. When ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). True positives were PTMs that were expected to be differential after accounting for changes in global protein abundance. True negatives were PTMs that were not expected to be differential after accounting for changes in global protein abundance. To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff. These results were then compared to the ground truth to determine TP and TN. Summary statistics were calculated, including the false discovery rate (FDR) (FP/(TP+FP)), recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). Finally, the summary statistics were compared across methods to analyze method performance.

For biological experiments where ground truth was not known, we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites. Comparisons were made before and after adjusting for changes in unmodified protein abundance and differences between the comparisons were noted. Each biological dataset presented a different experimental scenario, showing the applicability of the proposed method on a variety of experimental designs. All datasets in this paper were evaluated using the $MSstatsPTM$ package.

Further details reviewing the results of each experimental analysis can be seen in Supplementary Sec. \ref{supp-sec:experiments}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM corrected for high FDR in computer simulations}

The proposed approach was evaluated against two-sample $t$-test and Limma in computer simulations. 
In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in the unmodified protein resulted in a high FDR. After adjusting for changes in the unmodified protein, all methods performed similarly in terms of FDR. In a clean simulation, recall and accuracy of the proposed approach and Limma performed similarly, although Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the $t$-test method lagged behind the others. Two-sample $t$-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations. In contrast, the proposed approach and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM outperformed existing methods in simulations with missing values and low features}

In simulations with missing values and low features the proposed method outperformed $t$-test and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in Figure \ref{fig:sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than Limma's IQR and 10.4\% smaller than $t$-test's IQR. This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In a label-free benchmark experiment MSstatsPTM corrected for bias and outperformed existing methods}

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (Figure \ref{fig:spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's. While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance. Additionally, the background peptides, serving as the null model, show many false positives before adjustment. After adjustment the number of false positives decreased significantly. Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied. While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, using Limma resulted in many insignificant adjusted p-values. Using Limma would have resulted in missing the majority of differential PTMs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM corrected for bias in group comparison TMT experiment}

The results of this experiment are summarized in Figure \ref{fig:data4_plots}. Figure \ref{fig:data4_venn_diagram} shows the number of significant modified peptides before and after adjustment. More peptides became insignificant after adjustment than became significant. 3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues were corrected in adjustment, and the unconfounded abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.

In Figure \ref{fig:data4_profile_plot}, the modification of protein $GSDMD$ at site $K62$ showed the advantage of the proposed method. The modified peptide originally showed no abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This was contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted p-value was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change was much higher, $2.79$, and the adjusted p-value became significant, $5.25e-8$. In this case the effect of the modified peptide was strongly confounded with changes in the global protein. The proposed method allowed us to remove this confounding and estimate the true effect.

\subsubsection*{MSstatsPTM removed confounding in two mixture TMT experiment}

The results of this experiment are summarized in Figure \ref{fig:data5_plots}. Figure \ref{fig:data5_venn_diagram} shows the number of significant modified peptides before and after adjustment. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models. Again we tested if the decrease in significant peptides was due to removing confounding with changes in the umodified protein or increased variance after adjustment was applied. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied, 548 peptides became insignificant due to an increase in variance. This is only 3.4\% of the total peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

@@

\subsubsection*{In label-free experiment without a separate global profiling run, MSstatsPTM eliminated the bias due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}

As discussed in Section \ref{sec:exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted. Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model compared to other experiments, seen in Table \ref{tab:dataDescription}.

The results of this experiment are summarized in Figure \ref{fig:data6_plots}. After adjusting for changes in the unmodified protein, there were fewer significant modified peptides. In total, 2709 modified peptides became insignificant, while only 547 became significant. However, this was mainly due to not having a global profiling run, resulting in a lack of overlap between modified and unmodified peptides. In the bottom plot, only modified peptides that could be adjusted are shown. Here there were much fewer peptides that became insignificant after adjustment. 726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models. Only 25 PTMs became insignificant due to increased variance from adjustment.

>>>>

\medskip \noindent {\bf Experimental Design:}~\figref{benchmark-design}(a) overviews the experimental design. Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E. coli} lysate, and human spike-in ub-peptide mixture. Unmodified peptides from human lysate were viewed as the global proteome. Background {\it E. coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling between Mix 1 and 3, and between Mix 2 and 4.  50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides in a mixed background of lysates from {\it E. coli} and a human cell line. Quantitative changes in protein and site abundance of these 20 human proteins were the target of the benchmark. In particular, we distinguished the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed). The true log-fold changes between the relevant components of the relevant mixtures are summarized in~\figref{benchmark-design}(b).  Two replicate mixtures were created per condition.

@@

We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTM. The true positives were defined as PTMs with changes distinct from the overall changes in protein abundance. The true negatives were defined as PTM which, after accounting for the changes in the overall protein abundance, were not differentially abundant. Additional summaries were the false discovery rate $FDR=FP/(TP+FP)$, $Recall=TP/(TP + FN)$, and $Accuracy=(TP + TN) / (TP + TN + FP + FN)$.
For biological experiments with unknown ground truth, we compared the differentially abundant PTM  with and without adjusting for changes in unmodified protein abundance.

The evaluations were based on the $MSstatsPTM$ \todo{any details of the options? E.g., was Empirical Bayes moderation used in MSstatsPTM?} $MSstatsPTM$ was compared \todo{with methods XXX - why chosen? We did not compare to IsobarPTM because XXX. Do we say t-test or ANOVA?} All the evaluations were done at the adjusted p-value cutoff of $\alpha=.05$. More details are in Supplementary Sec. \ref{supp-sec:experiments}.

\todo{OV:Most of the following sections simply re-state the results that can already be read off the figures. The sections would have more value if in addition/instead of describing the figures, we could provide the reasons/insight into why we see this performance. Could you try? In general, no need to repeat what we see in the figure - this is what the figures are for.}

\todo{OV: Also, maybe combine two sections that make a similar point?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In clean simulations, MSstatsPTM corrected for high FDR of differentially abundant PTM}

\todo{OV: maybe say something about the fact that this is a group comparisons design that doesn't require many variance components, and as such is favorable to Limma and t-test? Later when you have repeated measurements, clarify that these situations are more challenging for the existing methods and therefore the difference in the results is more pronounced.}

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly. \todo{As expected when not using EB moderation in MSstatsPTM?,} Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the two-sample $t$-test method lagged behind. This is due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data. In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In noisy simulations, MSstatsPTM \todo{had a higher metric XXX than} the existing methods}

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and $Limma$. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In the label-free benchmark experiment, MSstatsPTM \todo{had a higher metric XXX than} the existing methods}

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than $Limma$ and 32.98\% smaller than $t$-test's. While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.

In~\figref{spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance. Additionally, the background peptides, serving as the null model, show many false positives before adjustment. After adjustment the number of false positives decreased significantly. Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied. While the proposed method and $Limma$ both correctly estimated the fold change of the spike-in peptides, using $Limma$ resulted in many insignificant adjusted p-values. Using $Limma$ would have resulted in missing the majority of differential PTMs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In the group comparison TMT experiment, MSstatsPTM corrected for bias \todo{OV: 'corrected for bias' is unclear. We talk about confounding. Maybe use a more specific term?}}

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} The results of this experiment are summarized in~\figref{data4_plots}. \figref{data4_venn_diagram} shows the number of significant modified peptides before and after adjustment. More peptides lost than gained differential abundance after the adjustment. 3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues were corrected in adjustment, and the unconfounded abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.

In~\figref{data4_profile_plot}, the modification of protein $GSDMD$ at site $K62$ showed the advantage of the proposed method. The modified peptide originally showed no abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This was contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted p-value was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change was much higher, $2.79$, and the adjusted p-value became significant, $5.25e-8$. In this case the effect of the modified peptide was strongly confounded with changes in the global protein. The proposed method allowed us to remove this confounding and estimate the true effect.

\subsubsection*{In the two-mixture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} The results of this experiment are summarized in Figure \ref{fig:data5_plots}. Figure \ref{fig:data5_venn_diagram} shows the number of significant modified peptides before and after adjustment. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models. Again we tested if the decrease in significant peptides was due to removing confounding with changes in the umodified protein or increased variance after adjustment was applied. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied, 548 peptides became insignificant due to an increase in variance. This is only 3.4\% of the total peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

@@

\subsubsection*{In label-free experiment without a separate global profiling run, MSstatsPTM eliminated the confounding due to changes in the unmodified protein, albeit less effectively than in the presence of a global profiling run}

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?}  As discussed in \secref{exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted. Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model compared to other experiments, seen in Table \ref{tab:dataDescription}.

The results of this experiment are summarized in~\figref{data6_plots}. After adjusting for changes in the unmodified protein, there were fewer significant modified peptides. In total, 2709 modified peptides became insignificant, while only 547 became significant. However, this was mainly due to not having a global profiling run, resulting in a lack of overlap between modified and unmodified peptides. In the bottom plot, only modified peptides that could be adjusted are shown. Here there were much fewer peptides that became insignificant after adjustment. 726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models. Only 25 PTMs became insignificant due to increased variance from adjustment.
